Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-03-27 RNARedPrint-Bioinfo2018.tex
Clarify potential mix-up
Minor changes

\hypersetup{
    bookmarks=false,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=true,        % show Acrobat’s toolbar?
    pdfmenubar=true,        % show Acrobat’s menu?
    pdffitwindow=true,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue!50!black,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=green!50!black           % color of external links
}

@@

\textcolor{red}{\bf RED ALERT! CHECK: IS THERE SOMETHING MISSING?}

>>>>

\hypersetup{
    bookmarks=false,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=true,        % show Acrobat’s toolbar?
    pdfmenubar=true,        % show Acrobat’s menu?
    pdffitwindow=true,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
    citecolor=red!70!black,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=red!70!black           % color of external links
}

@@

### 2018-03-27 supplement.tex
Minor changes

\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively. It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
\newcommand{\Parity}[1]{p_{#1}}
To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp. \Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.

\textcolor{red}{\textbf{SW: Introducing variables now conflicts with
    the specialization of constraint networks to RNA design
    networks. Luckily, there should be an even easier way to describe
    the complexity reduction without introducing new variables; ideally let's discuss}}

Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
\begin{enumerate}
\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
\end{enumerate}
Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition. Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}. Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph. Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd. It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.

The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node. It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}. Also, note that the number of nodes introduced by Step~\ref{step:intermediates} is bounded by $|CC|$, so that the introduction of these intermediate nodes does not impact the worst-case complexity. For the same reason, the complexity of the stochastic backtrack remains unchanged.

>>>>

\section{Exploiting constraint consistency to reduce the complexity}\label{sec:improvedComplexity}

\textcolor{red}{\textbf{SW: Simpler alternative to
    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
    with the (recent) specialization of constraint networks to RNA design
    networks.}}

While Alg.~\ref{alg:pf} computes messages values for
\emph{all} possible combinations of nucleotides for the positions 
% in the separator set, 
% yp: for all the positions in a node, right?
in a node, 
we observe here that many such combinations are \emph{not} required for computing all relevant
partition functions. In particular, the algorithm can be restricted to
consider only \Def{valid} combinations, satisfying the (hard)
constraints induced by valid base pairs.

RNA design introduces constraints due to the requirement of canonical
(aka complementary) base pairing, which can be exploited in a
particularly simple and effective way to reduce the complexity.  
As previously noted by \citet{Flamm2001}, the base pair complementarity induces a
bi-partition of each connected component within the base pair
dependency graph, such that the nucleotides assigned to the two set of
nodes in the partition are restricted to values in $\{\Ab,\Gb\}$ and
$\{\Cb,\Ub\}$ respectively. We call a partial sequence
\Def{cc-valid}, iff its determined positions are consistent with such a separation
for all determined positions of the same connected component.

% Let $CC=\{cc_1$, $cc_2, \dots\}$ denote the vertex sets of the
% connected components of the base pair dependency graph. In this graph,
% let each component have the bi-partition $cc_i=cc^0_i\cup cc^1_i$.

One can now modify Alg.~\ref{alg:pf}, on a tree decomposition
$(T,\chi)$, such that the message values
$\evalfor{\Message{u}{v}}{\val}$ are only computed for cc-valid
partial sequences $\val\in\partseqs(\separator{u}{v})$. Moreover, the
loop over $\val'\in\partseqs(\difference{u}{v})$ is restricted, such
that $\substitute{\val}{\val'}$ are cc-valid. Analogous restrictions are 
then implemented in the sampling algorithm Alg.~\ref{alg:sampling}.

The correctness of the modified algorithm still follows from an
induction argument: in this way, the message computation over one node
evaluates messages from its children only at cc-valid partial
sequences. The result of this computation is a message, which corresponds 
to the partition function, restricted to cc-valid partial sequences. 
Since invalid sequences have infinite energy, they do not contribute to 
the partition function, and the restricted partition function coincides 
with the initial one.

Remarkably, this restriction drastically improves the time
complexity. While the original algorithm enumerates all partial
sequences $\chi(v):=\substitute{\val}{\val'}$ such that
$\val\in\partseqs(\separator{u}{v})$ and
$\val'\in\partseqs(\difference{u}{v})$, the modified algorithm
only considers cc-valid sequences for $\chi(v)$ the nodes of $v$.
Namely,

For each connected component having at least a position $i$ in $\chi(v)$, 
the algorithm considers assigning any possible nucleotide $x\in \Sigma$ to $i$ and restricts
 the assignment of  for 
any remaining position $j$ of the same connected component, it restricts the domain of $j$ to base-pairs that are consistent with the parity .

Let $c(v)$ be the number of connected components involving positions
in $\chi(v)$; then, the first stage enumerates $2^{c(v)}$
combinations. For each such combination, the second stage enumerates
only $2^{|\chi(v)|}$ partial sequences, since there remain only two
possible nucleotide choices for each position. All message values for
node $v$ are thus computed in time
$\Theta(2^{c(v)+|\chi(v)|})$. Consequently, the total time complexity
of the modified algorithm is $$O(nk2^{\hat w}),$$ where $n$ is the RNA
length, $k$ is the number of structures, and
$\hat w := \max_{v\in T} c(v)+|\chi(v)|$.
%
From this directly follows the slightly looser complexity
bound $$O(nk2^{w+1}2^{c})$$ in terms of treewidth $w$ and the maximum
number $c$ of connected components represented in a node of the tree
decomposition.  Notably, $\hat w$, as well as $w+1+c$, can not exceed
$2(w+1)$, since $c\leq w+1$. Consequently the modified algorithm does
never increase the original complexity.

Finally, we remark that even stronger time savings could be possible
in practice, since cc-valid partial sequences can still violate
complementarity constraints, e.g. by assigning C and A to positions in
different sets of a partition, thus satisfying the bi-partition
constraints, where the positions base pair directly, rendering the
partial sequence invalid. Moreover, applications of the sampling
framework, can introduce additional constraints that further reduce
the number of valid partial subsequences. However, exploiting all such
constraints, in a complete and general way, likely causes significant
overhead in an implementation and does not straightforwardly improve
the theoretical bound.

%\section{Exploiting the bipartite nature of base pair dependency
%  graphs to reduce the complexity}\label{sec:first-improvedComplexity}
%\textcolor{red}{\textbf{TODO: discuss and merge with/replace by previous Section}}
%
%As already noted by \citet{Flamm2001}, the requirement of canonical
%base pairing induces a bi-partition of each connected component within
%the base pair dependency graph, such that the nucleotides assigned to
%the two set of nodes in the partition are restricted to values
%$\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
%%
%It follows that, within each connected component of the base pair
%dependency graph, the assignment of any position drastically restricts
%the domain of the remaining positions, \emph{i.e.} they may now be
%assigned to only 2 out of the original 4 nucleotides.
%\newcommand{\Parity}[1]{p_{#1}} To exploit this property, and further
%reduce the exponential factor of the complexity, we introduce the
%notion of \Def{parity} of a vertex/position, whereby vertices of
%\Def{even} (resp. \Def{odd}) polarities may only be assigned values in
%$\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must
%be even and vice versa, so the choice of a parity for any single
%vertex caracterizes the parity of all vertices in a bipartite
%connected subgraph.
%
%Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the base pair dependency graph, we consider a modified version of the cluster tree, obtained by:
%\begin{enumerate}
%\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
%\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
%\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
%\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
%\end{enumerate}
%Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition. Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}. Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the base pair dependency graph. Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd. It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.
%
%The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node. It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}. Also, note that the number of nodes introduced by Step~\ref{step:intermediates} is bounded by $|CC|$, so that the introduction of these intermediate nodes does not impact the worst-case complexity. For the same reason, the complexity of the stochastic backtrack remains unchanged.
